A Bayesian networks approach to infer social changes in burials in northeastern Taiwan during the European colonization period

### 2021-05-15 analysis/paper/paper.Rmd
justify the connection between theory and network expectation

<!--study social changes in pericolonial context using burial data as a proxy-->
In this paper, we go beyond ERGMs with a novel Bayesian approach that alleviates the computational issues and other limitations of ERGMs to enable clearer and more robust interpretations of the processes of network formation. Using burial data from an Iron Age site in northeastern Taiwan as a case study, we explore social changes by investigating the formation of material connections between burials. Social changes in Indigenous societies, when faced with colonial powers, are commonly observed in many parts of the world, especially European colonies where Indigenous economic, cultural, or socio-political aspects were substantially impacted [@Dietler2005; @Silliman2005; @Voss2005]. Recent studies demonstrate that the indirect effects of colonialism, or involvement in long-distance trade, may also have impacted Indigenous societies, this is known as a pericolonial context [@Acabado2017; @Trabert2017; @LWandBM2020ornament]. Burials are important for understanding past societies because material culture and biological records of burial behaviors can represent the social ranking of the deceased, and social relations between them [@Binford1971; @Drennan2010; @Saxe1970]. Burial treatments can further indicate social complexity or inequality where foreign goods from long-distance trade were used by Indigenous people to express status [@Carter2015; @Dolfini2019].

@@

The involvement in long-distance exchange is often associated with changes in political-economic strategies from a corporate mode to a network mode, where ambitious individuals are able to build their local power base and personal influence through the acquisition and distribution of foreign prestige goods [@Blanton1996; @Feinman2000; @Klehm2017]. Network-based societies are identified in some places in Asia that were invovled in long-distance trade [@Carter2015; @Ueda2016; @Liu2006]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans. Here we assume that trade goods could be used as prestige goods based on their archaeological contexts where very few of trade goods were found in middens and more in burials, along with the symbol as personal influence described in the Spanish records [@LiandWu2006]. In addition, the study of spatial distribution of trade ornaments in residential at Kiwulan presents a more clustered pattern after the European presence [@LWandBM2020ornament].

<!--theoretical framework and research questions-->
The features of network-based societies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks [@Blanton1996; @Feinman2000]. These may be contrasted with corporate-based societies that stress communal ritual elements, shared power, and where wealth differentiation, if any, would be associated with corporate groups, such as age groups, rather than individuals [@Siegel1999]. Using the theoretical framework of corporate-network strategies, we test a hypothesis that changes from a corporate mode (a structure showing more subgroups with less wealth differentiation) to a network mode (a centralized structure with more wealth differentiation) can be observed in the Kiwulan burial network after European arrival. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? and if so, (2) what are the major variables affecting or forming unequal social positions that might hint at social heterogeneity? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

<!-- our expectation based on hypothesis --> 
Our social change hypothesis is that differential access to foreign prestige goods after the European presence led to increased social inequality at Kiwulan, where the social structure changed from a more corporate strategy to a more network strategy [@Blanton1996; @Drennan2010; @Feinman2000]. This would be reflected by the pattern of intercorrelations of foreign prestige goods in burial networks. If social inequality gradually increased as we hypothesize, then we expect to observe a network structure with higher centralization (or popularity) and less transitivity (or clustering), along with network ties determined by wealth, indicative of a network-based society. In contrast, we predict the network before European arrival will show low centralization and high transitivity, and network ties associated with ritual elements, indicative of a corporate-based society. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shaped relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- theoretical framework for the results-->
A striking finding in our results is the change in network properties of Kiwulan burials from a more cohesive network with multiple subgroups toward a more centralized network with concentration of connections among fewer burials after the European arrival in the 17th century. This supports our pericolonial impact model that Kiwulan changed from a more corporate-based to a more networked-based society, as indicated in the grave goods and burial attributes. A corporate-based society, as we interpret the pre-European situation at Kiwulan, stresses shared power between individuals, communal rituals, and social inequality, if any, would be associated with groups [@Drennan2010; @Feinman2000; @Feinman2000political]. In contrast, a network-based society, as we interpret the post-European data from Kiwulan, presents wealth accumulation through individual networks, prestige goods manipulation, and trade monopolization [@Blanton1996]. It should be noted that the corporate-network continuum represents a dynamic process with different degrees of hierarchical complexity instead of static ideal-type stages [@Feinman2000].

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestigegoods. Our results indicate that the major factor contributing to burial networks for the pre-European network was ritual, and for the post-European network was wealth. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors were important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual could be a status indicator at the time before many foreign goods were introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences. Geographical proximity of burials, implying kinship association, has no effect on sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for both networks. We also note that no effects could be due to the unavailability of age (`r age_miss` out of `r pre_and_post` burials) and sex (`r sex_miss` out of `r pre_and_post` burials) values in our sample.

<!-- other evidence supporting the change in burial networks -->
We used burial data as a proxy to explore social relations reflected by interconnections of prestige goods based on assumptions that burials can represent social structures [@Binford1971; @Saxe1970], and individuals with similar prestige goods reflect similar access to trade, exchange, and gifting networks [eg. @Coward2013, pp. 252]. We argue that the social relations at Kiwulan changed because of contact with European colonial powers and associated foreign trade. This argument is supported by changes in burial network structures through time. It should be noted that manipulation of burial rituals by the living can cause a disconnect between a person’s status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982]. However, this issue can be reduced by comparing the results with evidence from other archaeological contexts, such as residential areas [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution when the Europeans were active in northern Taiwan that hints at an increased social differentiation [@LWandBM2020ornament]. This is consistent with the result of a more centralized burial network after the presence of Europeans, that supports a connection between a burial and the social role of the living person. Also, ethnographic records support our interpretation of burial goods as prestige goods or status items in the local Indigenous culture [@Borao2009; @LiandWu2006], so the interrelations represented by the flow of these goods observed in the burial data likely reflect social contexts of the living people at Kiwulan.

<!-- trade dynamic during this period and future work for comparison -->
We also recognize that trade dynamics between different Indigenous groups and other foreign presence, Han Chinese and Japanese, might have had contributed to social changes at Kiwulan. However, the current changes observed from burial networks and spatial distribution of trade items in the residential area at Kiwulan demonstrate consistent changes before and after the European arrival. Historical records also support the interaction and trade activities between Europeans and local Indigenous people. The trade items, including Anping jars, stonewares, Chinese porcelains, and beads found at Kiwulan were frequently found from archaeological sites with Dutch occupation, such as Fort Zeelandia. As a large village with continuous occupation since the 14th century in northeastern Taiwan, Kiwulan provides an ideal example to explore the reaction of Indigenous people to European colonial presence at a settlement scale. We still need more evidence from other sites in northeastern Taiwan to dive deeper into pericolonial effects, and compare evidence from the periphery with colonial bases, such as Heping Dao and Tamsui. However, Indigenous sites related to the 17th century with intensive excavation and sufficient evidence are rare to offer good comparison currently and future work is still needed.

>>>>

<!--study social changes in pericolonial context using burial data as a proxy-->
In this paper, we go beyond ERGMs with a novel Bayesian approach that alleviates the computational issues and other limitations of ERGMs to enable clearer and more robust interpretations of the processes of network formation. Using burial data from an Iron Age site in northeastern Taiwan as a case study, we explore social changes by investigating the formation of material connections between burials. Social changes in Indigenous societies, when faced with colonial powers, are commonly observed in many parts of the world, especially European colonies where Indigenous economic, cultural, or socio-political aspects were substantially impacted [@Dietler2005; @Silliman2005; @Voss2005]. Recent studies demonstrate that the indirect effects of colonialism, or involvement in long-distance trade, may also have impacted Indigenous societies, this is known as a pericolonial context [@Acabado2017; @Trabert2017; @LWandBM2020ornament]. Burials are important for understanding past societies because material culture and biological records of burial behaviors can represent the social ranking of the deceased, and social relations between them [@Binford1971; @Curet1998; @Drennan2010; @Saxe1970]. Burial treatments can further indicate social complexity or inequality where foreign goods from long-distance trade were used by Indigenous people to express status [@Carter2015; @Dolfini2019].

@@

The involvement in long-distance exchange is often associated with changes in political-economic strategies from a corporate mode to a network mode, where ambitious individuals are able to build their local power base and personal influence through the acquisition and distribution of foreign prestige goods [@Blanton1996; @Feinman2000; @Klehm2017]. Network-based societies are identified in some places in Asia that were involved in long-distance trade [@Carter2015; @Ueda2016; @Liu2006]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by European colonial activities in northern Taiwan. Foreign trade goods could be used as prestige goods based on their archaeological contexts where very few of trade goods were found in middens and more in burials, along with a symbol of personal influence described in the Spanish historical records [@LiandWu2006]. In addition, the study of spatial distribution of trade ornaments in residential area at Kiwulan demonstrates a more clustered pattern after the European presence, indicative of intentional accumulation of trade goods [@LWandBM2020ornament].

<!--theoretical framework and research questions-->
The features of network-based societies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks [@Blanton1996; @Feinman2000]. These are contrasted with corporate-based societies that stress communal ritual elements, shared power, and where wealth differentiation, if any, would be associated with corporate groups, such as age-based or kin-based, rather than individuals [@Siegel1999]. The European contact and associated trade activities could be the stimulus of competition for foreign resources among aggrandizers that led to social changes. Using the theoretical framework of corporate-network strategies, we test a hypothesis of the shift in organizational modes in relation to increased social inequality, from a more corporate to a more network society, after the European presence that can be observed in the Kiwulan burial network. We ask: (1) Do the burial network after European arrival present a more centralized structure with fewer subgroups compared to the burial network before European contact? and if so, (2) what are the major variables affecting or forming these two networks? Can we identify the shift in variables from ritual to wealth that associated with two different strategies? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

<!-- our expectation based on hypothesis --> 
Our social change hypothesis is that differential access to foreign prestige goods after the European presence led to increased social inequality at Kiwulan, where the social structure changed from a more corporate strategy to a more network strategy, presenting different organizational modes of societies [@Blanton1996; @Drennan2010; @Feinman2000]. Corporate-based societies stress shared power, multiple segments, and social cohesion through communal rituals, while network-based societies emphasize the accumulation of individual prestige, wealth, and resources by building personal networks. We used burial data as a proxy to explore social changes from a corporate strategy to a network strategy. This change can be identified though the pattern of intercorrelations of foreign prestige goods based on assumptions that burials can represent social structures [@Binford1971; @Saxe1970], and individuals with similar prestige goods reflect similar access to trade, exchange, and gifting networks [eg. @Coward2013, pp. 252].

If social inequality gradually increased after European arrival as we hypothesize, then we expect to observe a network structure with higher centralization and less transitivity, along with network ties determined by wealth, indicative of a network-based society. In contrast, we predict the network before European arrival will show low centralization and high transitivity, and network ties associated with ritual elements, indicative of a corporate-based society. High transitivity (or clustering) indicates multiple subgroups with shared burial goods, representing social cohesion instead of individual accumulation in an earlier time. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shaped relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- theoretical framework for the results-->
A striking finding in our results is the change in network properties of Kiwulan burials from a more cohesive network with multiple subgroups toward a more centralized network with concentration of connections among fewer burials after the European arrival in the 17th century. This supports our pericolonial impact model that Kiwulan changed from a more corporate-based to a more networked-based society, as indicated in the grave goods and burial attributes. Changes in mortuary practices could reflect the development of social or political complexity [eg. @Curet1998]. A corporate-based society, as we interpret the pre-European situation at Kiwulan, stresses shared power between individuals, communal rituals, and social inequality, if any, would be associated with groups [@Drennan2010; @Feinman2000; @Feinman2000political]. In contrast, a network-based society, as we interpret the post-European data from Kiwulan, presents wealth accumulation through individual networks, prestige goods manipulation, and trade monopolization [@Blanton1996]. It should be noted that the corporate-network continuum represents a dynamic process with different degrees of hierarchical complexity instead of static ideal-type stages [@Feinman2000].

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestige goods. Our results indicate that the major factor contributing to burial networks for the pre-European network was ritual, and for the post-European network was wealth. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors were important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual could be a status indicator at the time before many foreign goods were introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences, corresponding with the characteristics of corporate/network strategies. Geographical proximity of burials, implying kinship association, has no effect on sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for both networks. We also note that no effects could be due to the unavailability of age (`r age_miss` out of `r pre_and_post` burials) and sex (`r sex_miss` out of `r pre_and_post` burials) values in our sample.

<!-- other evidence supporting the change in burial networks -->
We argue that the social relations at Kiwulan changed because of contact with European colonial powers and associated foreign trade. This argument is supported by changes in burial network structures through time. It should be noted that manipulation of burial rituals by the living can cause a disconnect between a person’s status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982]. However, this issue can be reduced by comparing the results with evidence from other archaeological contexts, such as residential areas [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution when the Europeans were active in northern Taiwan that hints at an increased social differentiation [@LWandBM2020ornament]. This is consistent with the result of a more centralized burial network after the presence of Europeans, that supports a connection between a burial and the social role of the living person. Also, ethnographic records support our interpretation of burial goods as prestige goods or status items in the local Indigenous culture [@Borao2009; @LiandWu2006], so the interrelations represented by the flow of these goods observed in the burial data likely reflect social contexts of the living people at Kiwulan.

<!-- trade dynamic during this period and future work for comparison -->
We also recognize that trade dynamics between different Indigenous groups and other foreign presence, Han Chinese and Japanese, might have had contributed to social changes at Kiwulan. However, the current changes observed from burial networks and spatial distribution of trade items in the residential area at Kiwulan demonstrate consistent changes before and after the European arrival. Historical records also support the interaction and trade activities between Europeans and local Indigenous people. The trade items, including Anping jars, stonewares, Chinese porcelains, and beads found at Kiwulan were frequently found from archaeological sites with Dutch occupation, such as Fort Zeelandia. As a large village with continuous occupation since the 14th century in northeastern Taiwan, Kiwulan provides an ideal example to explore the reaction of Indigenous people to European colonial presence at a settlement scale. We still need more evidence from other sites in northeastern Taiwan to dive deeper into pericolonial effects, and compare evidence from the periphery with colonial bases, such as Heping Dao and Tamsui. Unfortunately, Indigenous sites related to the 17th century with intensive excavation and sufficient evidence as Kiwulan are rare to offer good comparison and future work is still needed.
